---
title: 'Day 3: Functions and Loops'
author: "María Ballesteros"
date: "2023-08-24"
runtime: slidy
output:
  slidy_presentation:
    df_print: paged
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

library(ggplot2)
library(tidyverse)
library(pacman)
```


# Check Point

+ On a scale of 1-10, how are you feeling about the material covered yesterday?
+ Any questions? 
+ Did you have anything good for dinner yesterday?
+ Did you do anything fun? 


# 20 minute ice breaker

You all wrote about your research interests in your statement of purpose, and did an amazing job (or else you would not be here)

+ What is one specific research question that you are excited to work on? 
+ What is one computational skill that you are hoping to learn to help answer that question? 

# Today's structue

+ Afternoon Session 1: Functions
+ Afternoon Session 2: Loops

# A new type of object: Lists

+  Lists are one of the most generic and flexible type of object. You can make an empty list by the function `list()`

```{r}
my_list <- list()
my_list
```
+ Think of lists sort of like collections of vectors and matrices

# Lists

+ And start filling it in. Slots on the list are invoked by double square brackets `[[]]`

my_list[[1]] <- "Chris"
my_list

+ each slot can be anything. 
  + What are we doing here? 
  + We are defining the 1st slot of the list `my_list` to be a vector `c(1, 2, 3, 4, 5)`
  
```{r}
my_list[[2]] <- 1:5
my_list

```


+ You can even make nested lists. Let's say we want the 1st slot of the list to be another list of three elements. 


```{r}
my_list[[2]][[1]] <- "R"
my_list
```

# Objects and classes

+ There are an infinite number of objects, because people make their own class of object. You can detect the type of the object (the class) by the function `class`

```{r}
class(my_list)  <- "Rstudio"
```
+ To change or create the class of any object, you can _assign_ it. To do this, assign the name of your class to character to an object's `class()`. 

+ We can start from a simple list. For example, say we wanted to store data about pokemon. 
+Because there is no pre-made package for this, we decide to make our own class. 


```{r}
 
pikachu <- list(name = "Pikachu",
                number = 25,
                type = "Electric",
                color = "Yellow")
pikachu
```

+ Like in base R, lists also have a specific "grammar"
  + One bracket calls the objects within the list
  + two brackets call specific components
```{r}
pikachu[2] # notice how it calling the entire "number" component
pikachu[[2]] #VS just the value 25

# To add more pokemon to the list, I can just replace the values in a component

pikachu[[1]] <- c("pikachu", "charmander")

# or

pikachu[[1]] <- c(pikachu[[1]],  "Eevee") # Since the output of pikachu[[1]]  is a vector, I can just add more to it
```


+ and we can give it any class name we want. 
class(pikachu) <- "pokemon"


# Functions

+ Objects can also have the class function.
  + All the functions that we use are objects that we load when we pull up libraries.
  + However, we can also create our own functions. 
  
Basic structure of a function  
```{r}
sum_1<-function(x){
  x+1
}

sum_1(x=4)

```

# lapply and sapply

+ We can also apply functions to multiple things at once:
  + To apply function accross a vector: lapply
  + To apply function accross a matrix: sapply
  
```{r}
all<- c(1:10)

lapply(all, sum_1)
```



